openTSNE: a modular \proglang{Python} library for t-SNE dimensionality reduction and embedding

### 2022-09-29 paper/main.tex
paper: update manuscript

Easy access to efficient implementations almost universally correlates with the widespread adoption of novel computational techniques. Despite the many theoretical advancements, most popular t-SNE libraries have been slow to incorporate them into their implementations. In this paper, we present \opentsne, our open-source \proglang{Python} implementation of t-SNE, complete with its recently proposed extensions. \opentsne is easy to install and includes precompiled binaries available through popular \proglang{Python} package managers. It provides a familiar API, making it suitable as a drop-in replacement for existing t-SNE implementations.

@@

To verify that that \opentsne was installed correctly, users can run \proglang{Python} and try to import the module by:
\begin{CodeChunk}
\begin{CodeInput}
>>> import openTSNE
>>> print(openTSNE.__version__)
0.6.2
\end{CodeInput}
\end{CodeChunk}

@@

@@

The results in this paper were obtained using \proglang{Python}~3.10.2 with the
\pkg{openTSNE}~v0.6.2 package. All results and figures can be reproduced using
the scripts found in the accompanying GitHub repository
\url{https://github.com/pavlin-policar/opentsne-paper}.

>>>>

Easy access to efficient implementations almost universally correlates with the widespread adoption of novel computational techniques. Despite the many theoretical advancements, most popular t-SNE libraries have been slow to incorporate them into their implementations.
In \proglang{Python}, the most commonly used implementation of the t-SNE algorithm comes from \pkg{scikit-learn}~\citep{pedregosa2011scikit}, which suffers from poor scalability. Other implementations from the \proglang{R}~\citep{team2013r,krijthe2015rtsne} and \proglang{Julia}~\citep{bezanson2017julia,julia_tsne} programming languages suffer from similar drawbacks. \pkg{MulticoreTSNE}~\citep{Ulyanov2016} enjoys better scalability but is likewise limited when applied to data sets containing millions of data points. Additionally, none of these implementations include recently proposed extensions that produce more globally consistent embeddings. \pkg{FIt-SNE}~\citep{linderman2019fast} alleviates these problems and provides bindings to both the \proglang{Python} and \proglang{R} programming languages. However, \pkg{FIt-SNE} is difficult to install and distribute and provides an API largely inconsistent with the \proglang{Python} data science ecosystem. None of these existing implementations provide solutions to adding new samples into existing embeddings.
To alleviate these problems, we present \opentsne, our open-source \proglang{Python} implementation of the t-SNE algorithm, complete with its recently proposed extensions. \opentsne is easy to install and includes precompiled binaries available through popular \proglang{Python} package managers. It provides a familiar API, making it suitable as a drop-in replacement for existing t-SNE implementations.

@@

To verify that that \opentsne was installed correctly, users can run \proglang{Python} and try to import the module by:
\begin{CodeChunk}
\begin{CodeInput}
>>> import openTSNE
>>> openTSNE.__version__
0.6.2
\end{CodeInput}
\end{CodeChunk}

@@

We provide three case studies demonstrating the usage of \opentsne in different settings using different combinations of hyperparameters to highlight different aspects of the given data sets.
The first case study highlights how varying the perplexity and resolution parameters can reveal different aspects of the underlying topology of the studied data set.
The second case study demonstrates how \opentsne may be used to create embeddings of massive data sets containing millions of data points and how its flexible API allows the quick and efficient construction of embeddings.
The final case study illustrates how we can use existing t-SNE embeddings to add new samples into the embedding space. This enables us to reuse carefully annotated visualizations for the rapid characterization of new, unseen data points.

@@

The results in this paper were obtained using \proglang{Python}~3.10.2 with the
\pkg{openTSNE}~v0.6.2 package. All results and figures can be reproduced using
the scripts found in the accompanying GitHub repository
\url{https://github.com/pavlin-policar/opentsne-paper}.
Detailed instructions are provided showing how to reproduce all materials and results shown in this manuscript.

All case studies can easily be reproduced on a consumer-grade laptop computer in minutes using the accompanying scripts. Note, however, that the second case study features a large data set and may require up to several hours to complete.

Scripts are included to reproduce the benchmarks of the different implementations of the t-SNE algorithm. However, please be aware that the full benchmark suite can take several days to complete due to the poor performance of other libraries. A smaller, illustratory benchmark suite is provided, more suitable for consumer-grade laptop computers, demonstrating the runtimes of the different t-SNE implementations. Please note, however, that the scaling advantages of \opentsne are diminished when running on smaller data sets and will not appear as significant as they would otherwise.
